Racial and Ethnic Differences in ETI: Outcomes

### 2024-07-22 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
ETI outcomes flowchart

```{r setup}
#| include: false
# Libraries
library(tidyverse)
library(lmerTest)
library(splines)
library(ggeffects)
library(gtsummary)
# Paths
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r filtering and flowchart}
# Variables for flowchart
n_enc_1 <- nrow(encounter)
n_people_1 <- length(unique(encounter$eDWID))
n_hosp_1 <- nrow(hospitalizations)
n_people_hosp_1 <- length(unique(hospitalizations$eDWID))
# Lung function cutoffs
lower <- 20
upper <- 150
# Remove impossible lung function values
low_fev1 = which(encounter$gli_fev1_ppred_rn)
encounter$gli_fev1_ppred_rn[encounter$gli_fev1_ppred_rn < 20 |
  encounter$gli_fev1_ppred_rn > 150] <- NA
encounter$gli_fvc_ppred_rn[encounter$gli_fvc_ppred_rn < 20 |
  encounter$gli_fvc_ppred_rn > 150] <- NA
encounter$gli_fev1fvc_ppred_rn[encounter$gli_fev1fvc_ppred_rn < 20 |
  encounter$gli_fev1fvc_ppred_rn > 150] <- NA
# Filter
encounter <- encounter %>%
  filter(
    !is.na(Days),
    Days >= -365.25 * 3 & Days <= 365.25 * 2,
    hospitalized == "No", !encounterlocation %in% c("Hospital", "Home IV"),
    pregnant != 1,
    !is.na(Modulator_trikafta_first_date)
  )
# Get some summary statistics by participant
t1_participant <- encounter %>%
  group_by(eDWID) %>%
  summarise(
    `Years With Data` = length(unique(reviewyear)),
    `Number of Encounters` = n(),
    `Number of Encounters Pre-ETI` = sum(Days <= 0),
    `Number of Encounters Post-ETI` = sum(Days > 0),
    Age = mean(encounterage, na.rm = T),
    across(all_of(continuous_outcomes), \(x) mean(x, na.rm = TRUE)),
    Race = names(sort(table(Race), decreasing = TRUE))[1],
    Insurance = names(sort(table(Insurance), decreasing = TRUE))[1],
    .groups = "drop"
  )
no_post <- t1_participant$eDWID[t1_participant$`Number of Encounters Post-ETI` == 0]
encounter <- encounter %>% filter(!eDWID %in% no_post)
t1_participant <- t1_participant %>% filter(!eDWID %in% no_post)
```

@@

1. Lung function percent predicted values < 20% or > 150% were excluded.
  - `r sum(encounter$gli_fev1_ppred_rn < 20,na.rm = T)` ppFEV1 values < 20% and `r sum(encounter$gli_fev1_ppred_rn > 150,na.rm = T)` over 150%.
2. Limited to 3 years before and 2 years after ETI start.
  - Based on `Modulator_trikafta_first_date` variable provided by CFFPR.
3. Hospitalizations and home IV episodes excluded.
  - Based on the care episodes dataset, there were `r length(unique(encounter$eDWID[encounter$hospitalized=="Yes"]))` unique participants with at least one care episode and a total of `r sum(encounter$hospitalized=="Yes")` encounters that were part of a care episode.
  - After filtering out these episodes, there were still 914 encounters with `encounterlocation` variable indicating either hospitalization or home IV, so these were also excluded.
4. Years of pregnancy excluded (we only have annualized pregnancy data, so don't know exact dates for this).
  - `r length(unique(encounter$eDWID[encounter$pregnant==1]))` unique participants with at least one care episode and a total of `r sum(encounter$pregnant==1, na.rm = T)` encounters during a pregnancy year.
5. Those missing `Modulator_trikafta_first_date` were automatically excluded from the models because they don't have a value for `Days`. This naturally means that people not eligible for Trikafta are not included in the models. 
6. 990 people with no encounters post-ETI were excluded.

```{r filtering}
# Additional quick cleaning steps

```

@@

# LOESS-smoothed outcome plots

```{r functions}
# Plots
loess_plot <- function(data = encounter, outcome, facet = F) {
  p <- ggplot(data, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    geom_smooth(method = "loess", se = F) +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
```

```{r results='asis'}
#| warning: false
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o))
}))
```

@@

## With Asian subgroup

```{r}
invisible(lapply(continuous_outcomes, function(o) {
  f <- as.formula(paste0(
    o,
    "~ age_group + Insurance + bs(Days, knots = 0, degree = 1) * Race + (1 | eDWID)"
  ))
  m <- lmer(f, data = encounter)
  predict_response(model = m, terms = c("Days[all]", "Race")) %>%
    plot() %>%
    print()
}))
```

## Without Asian subgroup (for clearer plotting)

>>>>

```{r setup}
#| include: false
# Libraries
library(tidyverse)
library(lmerTest)
library(splines)
library(ggeffects)
library(gtsummary)
library(ggflowchart)
# Paths
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

```{r filtering and flowchart}
# Variables for flowchart
n_enc_1 <- nrow(encounter)
n_people_1 <- length(unique(encounter$eDWID))
no_eti_date <- which(is.na(encounter$Modulator_trikafta_first_date))
n_people_no_eti_date <- length(unique(encounter$eDWID[no_eti_date]))
# Remove those without an ETI start date
encounter <- encounter[-no_eti_date, ]
n_enc_2 <- nrow(encounter)
n_people_2 <- length(unique(encounter$eDWID))
# Filter out hospitalizations
hosp <- which(encounter$hospitalized == "Yes" |
  encounter$encounterlocation %in% c("Hospital", "Home IV"))
n_people_hosp <- length(unique(encounter$eDWID[hosp]))
encounter <- encounter[-hosp, ]
n_enc_3 <- nrow(encounter)
n_people_3 <- length(unique(encounter$eDWID))
# Pregnancies
preg <- which(encounter$pregnant == 1)
n_people_preg <- length(unique(encounter$eDWID[preg]))
encounter <- encounter[-preg, ]
n_enc_4 <- nrow(encounter)
n_people_4 <- length(unique(encounter$eDWID))
# Days from ETI start
early <- which(encounter$Days < -365.25 * 3)
n_people_early <- length(unique(encounter$eDWID[early]))
late <- which(encounter$Days > 365.25 * 2)
n_people_late <- length(unique(encounter$eDWID[late]))
encounter <- encounter[-early, ]
encounter <- encounter[-late, ]
n_enc_5 <- nrow(encounter)
n_people_5 <- length(unique(encounter$eDWID))
# Get some summary statistics by participant
t1_participant <- encounter %>%
  group_by(eDWID) %>%
  summarise(
    `Years With Data` = length(unique(reviewyear)),
    `Number of Encounters` = n(),
    `Number of Encounters Pre-ETI` = sum(Days <= 0),
    `Number of Encounters Post-ETI` = sum(Days > 0),
    Age = mean(encounterage, na.rm = T),
    across(all_of(continuous_outcomes), \(x) mean(x, na.rm = TRUE)),
    Race = names(sort(table(Race), decreasing = TRUE))[1],
    Insurance = names(sort(table(Insurance), decreasing = TRUE))[1],
    .groups = "drop"
  )
no_post <- t1_participant$eDWID[t1_participant$`Number of Encounters Post-ETI` == 0]
no_post_visits <- which(encounter$eDWID %in% no_post)
encounter <- encounter[-no_post_visits, ]
n_enc_6 <- nrow(encounter)
n_people_6 <- length(unique(encounter$eDWID))
t1_participant <- t1_participant %>% filter(!eDWID %in% no_post)
# Lung function cutoffs
lower <- 20
upper <- 150
# Remove impossible lung function values
low_fev1 <- which(encounter$gli_fev1_ppred_rn < lower)
high_fev1 <- which(encounter$gli_fev1_ppred_rn > upper)
encounter$gli_fev1_ppred_rn[low_fev1] <- NA
encounter$gli_fev1_ppred_rn[high_fev1] <- NA

low_fvc <- which(encounter$gli_fvc_ppred_rn < lower)
high_fvc <- which(encounter$gli_fvc_ppred_rn > upper)
encounter$gli_fvc_ppred_rn[low_fvc] <- NA
encounter$gli_fvc_ppred_rn[high_fvc] <- NA
```

@@

```{r flowchart}
# Make labels
tots1 <- paste0(n_people_1, " People\n", n_enc_1, " Encounters")
no_date <- paste0(
  "No ETI Date:\n", n_people_no_eti_date, " People\n",
  length(no_eti_date), " Encounters"
)
tots2 <- paste0(n_people_2, " People\n", n_enc_2, " Encounters")
hosps <- paste0(
  "Care Episodes:\n", n_people_hosp, " People\n",
  length(hosp), " Encounters"
)
tots3 <- paste0(n_people_3, " People\n", n_enc_3, " Encounters")
pregs <- paste0(
  "Pregnancies:\n", n_people_preg, " People\n",
  length(preg), " Encounters"
)
tots4 <- paste0(n_people_4, " People\n", n_enc_4, " Encounters")
earlies <- paste0(
  "Visits > ", 365.25 * 3, " Days BEFORE ETI:\n", n_people_early, " People\n",
  length(early), " Encounters"
)
lates <- paste0(
  "Visits > ", 365.25 * 2, " Days AFTER ETI:\n", n_people_late, " People\n",
  length(late), " Encounters"
)
tots5 <- paste0(n_people_5, " People\n", n_enc_5, " Encounters")
no_posts <- paste0(
  "No Post-ETI Visits:\n", length(no_post), " People\n",
  length(no_post_visits), " Encounters (Pre-ETI)"
)
tots6 <- paste0(n_people_6, " People\n", n_enc_6, " Encounters")
fev1s <- paste0(
  length(low_fev1), " FEV1 < ", lower, "\n",
  length(high_fev1), " FEV1 > ", upper
)
fvcs <- paste0(
  length(low_fvc), " FVC < ", lower, "\n",
  length(high_fvc), " FVC > ", upper
)
# Assemble the chart!
flow_chart <- tibble(
  from = c(
    tots1, tots1, tots2, tots2, tots3, tots3, tots4, tots4, tots4, tots5,
    tots5, tots6, tots6
  ),
  to = c(
    tots2, no_date, tots3, hosps, tots4, pregs, tots5, earlies, lates, tots6,
    no_posts, fev1s, fvcs
  )
)
ggflowchart(flow_chart)
```

@@

<!-- # LOESS-smoothed outcome plots -->

<!-- ```{r functions} -->
<!-- # Plots -->
<!-- loess_plot <- function(data = encounter, outcome, facet = F) { -->
<!--   p <- ggplot(data, aes(x = Days, y = !!sym(outcome), color = Race)) + -->
<!--     geom_point( -->
<!--       shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F, -->
<!--       alpha = 0.1 -->
<!--     ) + -->
<!--     geom_smooth(method = "loess", se = F) + -->
<!--     theme_classic() -->
<!--   if (facet) { -->
<!--     p <- p + facet_wrap(~Race) -->
<!--   } -->
<!--   return(p) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r results='asis'} -->
<!-- #| warning: false -->
<!-- invisible(lapply(continuous_outcomes, function(o) { -->
<!--   cat("\n") -->
<!--   cat("\n") -->
<!--   cat("##", o) -->
<!--   cat("\n") -->
<!--   cat("\n") -->
<!--   print(loess_plot(outcome = o)) -->
<!-- })) -->
<!-- ``` -->

@@
